---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
## load the dplyr libary - I prefer this for grouping and summarizing
## load ggplot2 for the plots/graphs
library (dplyr)
library (ggplot2)

## this assumes you have downloaded, unzipped, and placed activity.csv into
## your working directory
df_orig <- read.csv("activity.csv")

head(df_orig)

## I had some issues with dates later, so this is ensuring the dates are of the date class
df_orig$date <- as.Date(df_orig$date , format="%m/%d/%Y")

## this dataframe will contain the original without the NAs
df_No_NAs <- na.omit(df_orig)

```

## What is mean total number of steps taken per day?

###Calculate the total number of steps taken per day

```{r}
## dplyr functions - create the group_by dataframe to use in the summarize function
byDates <- group_by(df_No_NAs, date)

## summarize to get the daily totals, 

dailyTotals <- summarize(byDates, sum=sum(steps))  

print(dailyTotals)

```

###Make a histogram of the total number of steps taken each day

```{r plot1 - histogram - dailyTotals}
##png("figure/plot 1 - dailyTotals.png")
p <- ggplot(data = dailyTotals, aes(x=date, y=sum)) + geom_bar(stat="identity")
print(p)
```

###Calculate and report the mean and median of the total number of steps taken per day

```{r}
## Summarize to get the mean and median of the daily totals
dailyMeanMedian <- summarize(dailyTotals, mean=mean(sum), median=median(sum)) 
print(dailyMeanMedian)

```

## What is the average daily activity pattern?

###Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r plot2 - time series - intervalMeans}
byInterval <- group_by(df_No_NAs, interval)
## summarize by the interval to get the means for each interval
intervalMeans <- summarize(byInterval, mean=mean(steps)) 
##print(intervalMeans)

## create the plot of interval means, one to a file, one to the document
##png("figure/plot 2 - intervalMeans.png")
q <- ggplot( data = intervalMeans, aes( interval, mean )) + geom_line() 
print(q)

```

###Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
maxInterval <- filter(intervalMeans, mean == max(mean))
print(maxInterval)
```

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
